Chapter 02: Our first model,Grand Old Logistic Regression

### 2020-09-20 DATA622/M02-glm.rmd
Minor bugfix

```{r}
names(heart)[[1]]<-'age'
```
Always trust but verify that our indexing worked...
```{r}
names(heart)
```
# Are the covariates correlated?
Numerically review the pairwise correlation amongst independent variables
```{r}
(corheart<-cor(heart[,1:13]))
```
# Are there any constant feature vectors?
Let us find out if there are any constant variables.
Constant variables cannot possibly contribute to predicting the target.
For example, given this simple birds dataset, and that I saw a bird with two legs,
can you identify which bird that might be?
```{r}
birds<-data.frame(legs=rep(2,8),name=c('crow','pigeon','owl','eagle','sparrow',
'duck', 'woodpecker', 'humming bird'),stringsAsFactors=F)
birds
```
Let us write a utility function we can call on any list...
we can also do length(unique(x)) < 1
```{r}
isConstant<-function(x) length(names(table(x)))<2
apply(heart,2,isConstant)
```
Let us plot,correlations using pairs function
```{r}
library(plotly)
p <- pairs(heart)
```
## What is our class/label variable?
Is the problem at hand *binary classification* or *multi-class classification!*?
Let us assume, our customers want us to predict **target** given the rest.
```{r}
classLabels<-table(heart$target) # this is the actual class distribution both test/train partitions must be close to this.
print(classLabels)
names(classLabels)
```
# Class Imbalance
Let us try to understand the *distribution of the target variable.* Let us make sure
the distribution is balanced. 
### When is a class imbalanced?
If we find 1(minority) in 10 in the extreme case
or anything less than 1 (minority) in 6, we will consider our dataset to be imbalanced.
These are not hard rules but my intuition is as follows
if we are going to create training data with 70/30 split then a random 
sample that contains none or very few of the minority class becomes likely.
Our efforts to train a model is not going to be very fruitful. The model
may not know how to detect minority class or overwhelmed by majority class signature.

@@

# File Drawer Problem
Include all results from all the experiments, regardless of the results whether it is in agreement with your expectation or NOT. Eliminate the so called File Drawer problem -- of not including unfavorable results --.
```{r}
coef_summary_glm_model
coef_summary_glm_model[,4]<0.05
row_names<-row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,])
row_names
summary_glm_model$aic
summary_glm_model$null.deviance
summary_glm_model$deviance
ifelse(summary_glm_model$deviance<summary_glm_model$null.deviance,"model has improved","model has not helped")
```
### p-values
Now let us rerun the model with the signifcant variables, that is, with p-value <0.05
```{r}
formulastr<-paste('target~',paste(row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,]),collapse='+'),sep='')
formulastr
paste(row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,]),collapse='+')
model2<-glm(formulastr,data=heart,family='binomial')
summ_model2<-summary(model2)
coef_summ_model2<-coef(summ_model2)
coef_summ_model2
summ_model2$aic
summ_model2$null.deviance
summ_model2$deviance
summary_glm_model$deviance
ifelse(summary_glm_model$deviance<summary_glm_model$null.deviance,"model has improved","model has not helped")
```
So we removed predictors based on p-value and our deviance of the
revised model is higher by about 6%.
So let us compute the VIF factors for the model predictors and 
determine if we removed some variables that should not have been removed.
```{r}
if(!require(car))install.packages('car')
library(car)
vif_model<-vif(glm_model)
```
VIF close to 1 indicates the variables are not correlated
Above 4 is highly correlated.
```{r}
vif_model
vif_model[vif_model>4]
names(vif_model)
nl<-names(vif_model[vif_model<4])
(newformulastr<-paste('target~',paste(nl,collapse='+')))
```
We learned from the entire dataset. Let us get the target variable into a vector.
```{r}
Y<-heart[[14]]
head(Y)
```
Now let us predict using our model. The model has already seen the data.
We do not expect our model, to reproduce at 100%. But we do expect that our model performs better than random guess. If our model under-performs random guess, then that would be an indication of the model's inability to learn at all.
Here in such cases that model is deficient and is a case of under-fitting.
```{r}
newmodel<-glm(formulastr,data=heart,family='binomial')
summary(newmodel)
summnewmodel<-summary(newmodel)
(summnewmodel$aic)
(summnewmodel$deviance)
(summnewmodel$null.deviance)
(p_values<-coef(summnewmodel)[,4])
```
Now let us compute the performance of the classifier.As indicated above,
a random choice would have scored 0.5. If our model is any good it must perform
better than 0.5.
```{r}
table(p_values<0.05)
predYprob<-predict(newmodel,heart[,1:13],type='response')
predY<-ifelse(predYprob<0.5,0,1)
table(predY)
table(heart[[14]])
table(heart[[14]],predY)
library(caret)
require(caret)
cfm<-caret::confusionMatrix(table(heart[[14]],predY))
cfm
```
Accuracy is 84.49%.
But this is from data the model has already seen.

>>>>

```{r}
names(heart)[[1]]<-'age'
```
Always trust but verify that our indexing worked...
```{r}
names(heart)
```
# Are the covariates correlated?
Numerically review the pairwise correlation amongst independent variables
```{r}
(corheart<-cor(heart[,1:13]))
```
# Are there any constant feature vectors?
Let us find out if there are any constant variables.
Constant variables cannot possibly contribute to predicting the target.
For example, given this simple birds dataset, and that I saw a bird with two legs,
can you identify which bird that might be?
```{r}
birds<-data.frame(legs=rep(2,8),name=c('crow','pigeon','owl','eagle','sparrow',
'duck', 'woodpecker', 'humming bird'),stringsAsFactors=F)
birds
```
Let us write a utility function we can call on any list...
we can also do length(unique(x)) < 1
```{r}
isConstant<-function(x) length(names(table(x)))<2
apply(birds,2,isConstant)
```
```{r}
isConstant<-function(x) length(names(table(x)))<2
apply(heart,2,isConstant)
```
Let us plot,correlations using pairs function
```{r}
library(plotly)
p <- pairs(heart)
```
## What is our class/label variable?
Is the problem at hand *binary classification* or *multi-class classification!*?
Let us assume, our customers want us to predict **target** given the rest.
```{r}
classLabels<-table(heart$target) # this is the actual class distribution both test/train partitions must be close to this.
print(classLabels)
names(classLabels)
```
# Class Imbalance
Let us try to understand the *distribution of the target variable.* Let us make sure
the distribution is balanced. 
### When is a class imbalanced?
If we find 1(minority) in 10 in the extreme case
or anything less than 1 (minority) in 6, we will consider our dataset to be imbalanced.
These are not hard rules but my intuition is as follows
if we are going to create training data with 70/30 split then a random 
sample that contains none or very few of the minority class becomes likely.
Our efforts to train a model is not going to be very fruitful. The model
may not know how to detect minority class or overwhelmed by majority class signature.

@@

# File Drawer Problem
Include all results from all the experiments, regardless of the results whether it is in agreement with your expectation or NOT. Eliminate the so called File Drawer problem -- of not including unfavorable results --.
```{r}
coef_summary_glm_model
coef_summary_glm_model[,4]<0.05
row_names<-row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,])
row_names
summary_glm_model$aic
summary_glm_model$null.deviance
summary_glm_model$deviance
ifelse(summary_glm_model$deviance<summary_glm_model$null.deviance,"model has improved","model has not helped")
```
### p-values
Now let us rerun the model with the signifcant variables, that is, with p-value <0.05
```{r}
formulastr<-paste('target~',paste(row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,]),collapse='+'),sep='')
formulastr
paste(row.names(coef_summary_glm_model[coef_summary_glm_model[,4]<0.05,]),collapse='+')
model2<-glm(formulastr,data=heart,family='binomial')
summ_model2<-summary(model2)
coef_summ_model2<-coef(summ_model2)
coef_summ_model2
summ_model2$aic
summ_model2$null.deviance
summ_model2$deviance
summary_glm_model$deviance
ifelse(summary_glm_model$deviance<summary_glm_model$null.deviance,"model has improved","model has not helped")
```
So we removed predictors based on p-value and our deviance of the
revised model is higher by about 6%.
So let us compute the VIF factors for the model predictors and 
determine if we removed some variables that should not have been removed.
```{r}
if(!require(car))install.packages('car')
library(car)
vif_model<-vif(glm_model)
```
VIF close to 1 indicates the variables are not correlated
Above 4 is highly correlated.
```{r}
vif_model
vif_model[vif_model>4]
names(vif_model)
nl<-names(vif_model[vif_model<4])
(newformulastr<-paste('target~',paste(nl,collapse='+')))
```
We learned from the entire dataset. Let us get the target variable into a vector.
```{r}
Y<-heart[[14]]
head(Y)
```
Now let us predict using our model. The model has already seen the data.
We do not expect our model, to reproduce at 100%. But we do expect that our model performs better than random guess. If our model under-performs random guess, then that would be an indication of the model's inability to learn at all.
Here in such cases that model is deficient and is a case of under-fitting.
```{r}
newmodel<-glm(newformulastr,data=heart,family='binomial')
summary(newmodel)
summnewmodel<-summary(newmodel)
(summnewmodel$aic)
(summnewmodel$deviance)
(summnewmodel$null.deviance)
(p_values<-coef(summnewmodel)[,4])
```
Now let us compute the performance of the classifier.As indicated above,
a random choice would have scored 0.5. If our model is any good it must perform
better than 0.5.
```{r}
table(p_values<0.05)
predYprob<-predict(newmodel,heart[,1:13],type='response')
predY<-ifelse(predYprob<0.5,0,1)
table(predY)
table(heart[[14]])
table(heart[[14]],predY)
library(caret)
require(caret)
cfm<-caret::confusionMatrix(table(heart[[14]],predY))
cfm
```
Accuracy is 84.49%.
But this is from data the model has already seen.
